---
title: "Outlying mean index"
author: "Elke Hendrix"
date: "February 28, 2019"
output: html_document
---

Load the required libraries:

```{r load_libs}
library(adehabitatHS, quietly = T)
library(raster, quietly = T)
library(SDMTools, quietly = T)
library(factoextra, quietly = T)
library(ecospat, quietly = T)
library(cluster, quietly = T)
```

Set the location of the root directory of the git repository
```{r globals}
REPO_HOME <- paste(getwd(),'/../',sep='')
```

Import the taxa list 

```{r taxa}
taxa.names <- scan(paste(REPO_HOME, "/data/filtered/taxa.txt", sep = ""), sep = "\n", what = character())
```

Open the abiotic environmental raster layers in the GIS repository

```{r layers}

currentEnv1 = raster::getData(
  "worldclim", 
  var="bio", 
  res=5, 
  path=paste(REPO_HOME, '/data/GIS', sep=''), 
  download=T
)

# The location of the TIFF file with the stacked layers described directly above
files_list <- list.files(paste(REPO_HOME,'/data/GIS/5_deg', sep=''))
additional_datasets<- stack()
for (i in 1:length(files_list) ){
# Open the raster file in the list
directorypath<- paste(REPO_HOME,'/data/GIS/5_deg/', files_list[i], sep='')
name<- files_list[i]
dataset<- raster(directorypath)
additional_datasets<-stack(additional_datasets,dataset)
}

names(additional_datasets)<- c("Aspect",
  "BulkDensity",
  "ClayPercentage",
  "annualPET",
  "aridityIndexThornthwaite",
  "climaticMoistureIndex",
  "continentality",
  "embergerQ",
  "growingDegDays0",
  "growingDegDays5",
  "maxTempColdest",
  "minTempWarmest",
  "monthCountByTemp10",
  "PETColdestQuarter",
  "PETDriestQuarter",
  "PETseasonality",
  "PETWarmestQuarter",
  "PETWettestQuarter",
  "thermicityIndex",
  "OrganicCarbon",
  "PhCaCL",
  "slope")

currentEnv <- stack(additional_datasets, currentEnv1)

```

The raster layers are transformed to a spatialpixelsdataframe

```{r taxa}
rtoAsc<- as(currentEnv, "SpatialPixelsDataFrame")
proj4string(rtoAsc) <- CRS("+proj=longlat +datum=WGS84")
```


Transform the csv files with longitude and latitude values to a SpatialPointsDataFrame

```{r SpatialPointDataFrame}
# create an empty SpatialPointDataFrame to populate in the following loop

spdf<-new("SpatialPointsDataFrame", coords = structure(numeric(0), .Dim = c(0L, 2L),                  .Dimnames = list(NULL, c("x", "y"))),  
    bbox = structure(c(1, 1, 1, 1), .Dim = c(2L, 2L),                         
                     .Dimnames = list(c("x", "y"),
                                      c("min", "max"))),
    proj4string = new("CRS", projargs = "+proj=longlat +datum=WGS84")) 

# populate the empty dataframe with longitude and latitude values from the taxa list
#c(1:134, 136:150, 152:154)
for (i in c(101:134, 136:150, 152:154)){
  valid.model.prediction <- sprintf("%s/results/per_species/%s/valid_maxent_prediction.rda", REPO_HOME, taxa.names[i])
    env <- new.env()
    nm <- load(valid.model.prediction, envir = env)[[1]]
    raster<- env[[nm]]
    raster[raster == 0] <- NA
    points<-rasterToPoints(raster, spatial=FALSE)
    points.number<- nrow(points)
    name<- paste(taxa.names[i])

    spdf2<- SpatialPointsDataFrame(points[,-3], data.frame(species = rep(name, points.number)))
  proj4string(spdf2) <- CRS("+proj=longlat +datum=WGS84")
  spdf<- rbind(spdf, spdf2)
}

```

Two dataframes are constructed:
1. a dataframe (dfavail) with all the values in the SpatialPixelsDataframe i.e. the traits 
2. a dataframe (dfused) with the amount of occurences per pixel  


```{r create_dataframes}
cp <- count.points(spdf, rtoAsc)

dfavail <- slot(rtoAsc, "data")
dfavail <- tibble::rowid_to_column(dfavail, "ID")
dfavail<- na.omit(dfavail)

dfused <- slot(cp, "data")
dfused <- tibble::rowid_to_column(dfused, "ID")

dfused<- subset(dfused, dfused$ID %in% dfavail$ID )
dfused <- subset(dfused, select = -ID )
```

A PCA analysis is conducted to standardize all the environmental variables. This standardized dataframe is used a an input dataset for the niche function to calculate the OMI per species. The OMI is used to plot both niche breadth and niche position per species. 

```{r PCA_analysis}
dud<- dudi.pca(dfavail[2:42], scannf=FALSE)
```


```{r OMI_analysis}
nic<- niche(dud, dfused, scannf=FALSE)

## This dataframe shows the averages per species for the standardized environmental variables
p<-nic$tab

normalized.csv<- paste(REPO_HOME, "/Results/OMI/normalized_values.csv", sep="")
write.csv(p, normalized.csv)
```

calculate gower's distance

```{r gowers_distance}
# load the normalized values from the github
normalized.repo <- paste(REPO_HOME, 'results/OMI/normalized_values.csv', sep = '')
normalized.values<- read.csv(normalized.repo)

# set names of species to rownames
tree.withouthnames<-normalized.values[,-1]
rownames(tree.withouthnames) <- normalized.values[,1]

# save gowers dataframe 
overlap.csv<- paste(REPO_HOME, "/Results/maxent/gowers_overlap.csv", sep="")
gow<- daisy(tree.withouthnames[2:42], metric ="gower", stand=FALSE)
gow_df<- as.matrix(gow)

  if (!file.exists(overlap.csv)) {
write.csv(gow_df, overlap.csv)
  }

# create tree
# compute distances, i.e. inverse of overlap
distances <- as.dist(gow_df)

hclust.tree <- hclust(distances)
tree <- as.phylo(hclust.tree)


```


Now we can assess whether domesticated Ungulates come from specific areas in niche space:

```{r resample_niche_clustering}

# vector of all tips
dom.taxa <- c(
  "Bos_frontalis_gaurus", "Bos_grunniens_mutus", "Bos_javanicus", "Bos_taurus_primigenius",
  "Bubalus_bubalis_arnee", "Camelus_bactrianus", "Camelus_dromedarius", "Capra_hircus_aegagrus",
  "Equus_przewalskii", "Lama_glama_guanicoe", "Ovis_aries_orientalis",
  "Rangifer_tarandus", "Sus_scrofa", "Vicagna_vicugna"
)

# tips in tree
dom.tips <- tree$tip.label[ which(tree$tip.label %in% dom.taxa) ]

# this function calculates the average pairwise patristic distance among an
# input vector of tip labels
mean_pw_distance <- function(tree, tip.vector) {

  # calculate size of distance matrix, instantiate vector
  nt <- length(tip.vector)
  nd <- ((nt * nt) - nt) / 2
  dist.vector <- vector(mode = "numeric", length = nd)

  # do all pairwise comparisons
  k <- 1
  max <- nt - 1
  for (i in 1:max) {
    min <- i + 1
    for (j in min:nt) {
      dist.vector[k] <- phytools::fastDist(tree, tip.vector[i], tip.vector[j])
      k <- k + 1
    }
  }
  return(mean(dist.vector))
}

# calculate average pairwise distance (APD) among domesticates
dom.dist <- mean_pw_distance(tree, dom.tips)

# take random tip samples, compute APD among these many time
sam.num <- 1000
sam.dist <- vector(mode = "numeric", length = sam.num)
for (i in 1:sam.num) {
  sam.tips <- sample(tree$tip.label, length(dom.tips), replace = F)
  sam.dist[i] <- mean_pw_distance(tree, sam.tips)
}


{
  clustering.pdf.file <- sprintf("%s/results/maxent/clustering3.pdf", REPO_HOME)
  pdf(file = clustering.pdf.file)
  h <- hist(
    sam.dist,
    xlab = sprintf("Mean pairwise patristic distance in random sample of %i tips (n=%i)", length(dom.tips), sam.num),
    col = "lightblue",
    main = "Domesticated Ungulates cluster in niche space"
  )

  # fit normal distribution
  xfit <- seq(min(sam.dist), max(sam.dist), length = length(sam.dist))
  yfit <- dnorm(xfit, mean = mean(sam.dist), sd = sd(sam.dist))
  yfit <- yfit * diff(h$mids[1:2]) * length(sam.dist)
  lines(xfit, yfit, col = "black", lwd = 2)

  # draw ± 2 * standard deviation
  abline(v = mean(sam.dist) + 2 * sd(sam.dist), col = "green")
  abline(v = mean(sam.dist) - 2 * sd(sam.dist), col = "green")

  # indicate domesticated APD
  abline(v = dom.dist, col = "red")
  dev.off()
}
```
